---
title: "Super Tricky Challenge"
author: "Tim Riffe"
date: "July 3, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The super tricky challenge

1. start with HFC (use those groups!)
2. Get TFR and MAB, but keep the age detail (`mutate()`)
3. Make TFR intervals, for example rounded to .25.
4. Make a column that identifies the highest and lowest MAB (like, maybe it uses NA for the others).
5. Within each interval pick out the lowest observed and highest observed MAB, and throw out the rest (`filter()`.
6. install the `ggridges` package.
7. make a ridge plot where TFR intervals are the levels, and on each plot two distributions: the distribution (ASFR) belonging to the highest and lowest observed MAB.

First: I copy-paste the data prep we used before:

```{r, fig.height = 10, fig.width = 5}
library(tidyverse)
run_this <- FALSE
if (run_this){
HFC <- read_csv("HFC_ASFRstand_TOT.txt", na = '.')

HFC <- HFC %>%
	mutate(YearInt = Year2 - Year1 + 1) 
HFC$sub_id <- group_indices(HFC,
							Country, 
						    Year1, 
							AgeDef, 
							Vitality, 
							Collection, 
							SourceType, 
							RefCode,
							YearInt)

HFCextremes <- HFC %>% 
	group_by(sub_id) %>% 
	mutate(TFR = sum(ASFR, na.rm = TRUE),
		   MAB = sum(ASFR * Age, na.rm = TRUE) / TFR,
		   TFRint = TFR - TFR %% .25) %>% 
	ungroup() %>% 
	group_by(TFRint) %>% 
	mutate(MABmax = MAB == max(MAB),
		   MABmin = MAB == min(MAB)) %>% 
	filter((MABmax | MABmin) & (TFR < 6)) %>% 
	mutate(extremes = ifelse(MABmax,"high MAB", "low MAB"))

saveRDS(HFCextremes, file = "HFCextremes.rds")
}

HFCextremes <- readRDS("HFCextremes.rds")
library(ggridges)

ggplot(HFCextremes, mapping = aes(x = Age, 
								  y = as.factor(TFRint))) +
	geom_ridgeline(aes(height = ASFR, fill = extremes), 
				   scale = 4, alpha = .5)

```

