---
title: "Monday Notes"
author: "Tim Riffe"
date: "July 1, 2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What is `R`?

1. It's functional:
```{r}
a <- 43
b <- 21
a+b
# same thing but this actually looks like a function:
sum(a,b)
```

2. extensible: people make packages

```{r, eval = FALSE}
install.packages("gapminder")
# install.packages("adiv")
install.packages("tidyverse")
```

3. a statistical language

```{r}
x <- rnorm(10)
y <- x * 2 * runif(10, min = .9, max = 1.1)
lm(y~x)
```

# gapminder

Gapminder is a tidy dataset.

Right about here I showed examples of tidy and
not tidy datasets.
```{r}
library(gapminder)
head(gapminder)

```
We also see code to look into a data object and see what's there. Plus a clicky way.

And then Kristen asked about case sensitivity in `R` (yes it is!) and we invoked an error and then I showed how to search for errors. The `searcher` package seems pretty great.

See, try typing this in the console (both lines)
```{r, eval = FALSE}
library(searcher)
str(Gapminder) # wrong case!
search_google() # what happened?
"a" + 1
search_stackoverflow()
```
It would appear that if you want to use this search shortcut, then it's better to copy and paste the error straight in the search function as a character string, e.g.: `search_stackoverflow("Error in "a" + 1 : non-numeric argument to binary operator")`. 

FYI: if you knit this document with `install.packages()` or anything trying to use the internet then it can hang. Best to include `eval = FALSE` in the chunk statement.


```{r}
library(tidyverse)
p <- ggplot(data = gapminder,
			mapping = aes(x = gdpPercap,
					      y = lifeExp,
						  color = continent)) 
p <- p + geom_point() 
p
```

Now we can *add* elements/metadata/instructions to the plot to make it the way we want:

```{r}
p <- ggplot(data = gapminder,
			mapping = aes(x = gdpPercap,
						  y = lifeExp))
p + 
	geom_point(alpha = .3, 
			   size = .8,
			   mapping=aes(color = continent)) +
	scale_x_log10() +
	geom_smooth(method = "loess")
```

Keys to find:
```{r}
# <- ^ * + - {} () [] ` ~
```

```{r}
p <- ggplot(gapminder,
			mapping = aes(x = gdpPercap,
						  y = lifeExp))
p + 
    geom_point() + 
	facet_wrap(~continent) +
	scale_x_log10() + 
	geom_smooth(method = "lm")
```

```{r}
p <- ggplot(gapminder,
			mapping = aes(x = gdpPercap,
						  y = lifeExp,
						  color = continent))
p + 
	geom_point() + 
	facet_wrap(~year) + 
	scale_x_log10() 

```

# Exercise 1

Smoothers: Do a Preston scatterplot panel by continent, showing different smoothing methods on top of the point clouds. For your own thinking: how come different methods gives such different uncertainties and what does that say about the data vs the method (way out of the realm of this course)?

```{r}
p <- ggplot(gapminder,
			mapping = aes(x = gdpPercap,
						  y = lifeExp)) +
	 geom_point(alpha = .3) + 
	facet_wrap(~continent) +
	scale_x_log10()
# 1) linear fits
p + geom_smooth(method = 'lm')
# 2) loess fits
p + geom_smooth(method = "loess")
# 3) GAM
p + geom_smooth(method = "gam", 
				formula = y~poly(x, 2))
# 4) GLM
p + geom_smooth(method = "glm")

```

# Exercise 2

There is a geom called `geom_density()` that draws a density function fit to your data. It only wants an `x`, not a `y`. You can have different overlapped densities by specifying `alpha = .5`, `fill=x` (where x is the thing you want to vary color on, continent in our case). You can subset data using `filter(gapminder,continent != "Oceania")`, to get rid of Oceania for example. Your challenge: Make a two panel plot, showing the `gdpPercap` density by continents in 1952 and 2007, where the density for each continent has a different semitransparent color. This will exercise like 5 things we learned, plus you'll use subsetting. This one's for Mr. Rosling. Do the same for `lifeExp`.

```{r}
gap2 <- filter(gapminder, continent != "Oceania" &
			   	((year == 1952) | (year == 2007))
			   )
gap2 <- filter(gapminder, continent != "Oceania" &
			   	year %in% c(1952, 2007)
			   )
ggplot(data = gap2,
	   mapping = aes(x = gdpPercap,
	   			  fill = continent)) +
	geom_density(alpha = .5) + 
	facet_wrap(~year) + 
	scale_x_log10()

ggplot(data = gap2,
	   mapping = aes(x = lifeExp,
	   			  fill = continent)) +
	geom_density(alpha = .5) + 
	facet_wrap(~year) 
```





